Unsupervised training of maximum-entropy models for lexical selection in rule-based machine translation

### 2015-02-09 2015-eamt-melsa/2015-eamt-melsa.tex
comment out worked example... one page less
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@58791 72bbbca6-d526-0410-a7d9-f06f51895060

As our method is unsupervised, and therefore the occurrences of
specific lexical selection events \((s,t,c)\) cannot be counted, a
target language model \(P_{TL}(\cdot)\) is used to compute a value for
the fractional count for disambiguation path \(g_i\), $p(g_i|S)$ after
suitable normalisation:
\begin{equation}
 \label{eq:normalising}
   p(g_i|s) = \frac{P_{TL}(\tau(g_i,S))}{\sum_{g_i \in G} P_{TL}(\tau(g_i,S))}
\end{equation}
~\\
The maximum-entropy model is trained instead using the fractional count
$p(g_i|S)$ for the events $(s,t,c)$ found in \(g_i\), that is, when in
\(g_i\) $t$ is the translation for \(s\) in context \(c\). That is, as
if event $(s,t,c)$ had been seen a fractional number $p(g_i,s)$ of
times. We prune features occurring less than a certain number of times 
in the corpus, using a development corpus to guide pruning (see section~\ref{sec:eval-ref-results}).
%@@@ \comment{Reword this to avoid repetition} 
The method used here
for lexical selection is analogous to the method used by
\cite{sanchez08b} to train a hidden-Markov-model-based part-of-speech
tagger in a rule-based machine translation system. 
Table~\ref{tb:workedexample} shows how a collection of samples
\(\mathcal{G}\) for the translation of Basque sentences containing the
word \emph{arrain}.\footnote{The word \emph{arrain} `fish' has two translations
in Spanish, the first \emph{pez} refers to the living animal, whereas
\emph{pescado} refers to the animal as food.} 
%\comment{FSM: The correct translations
%  should be clearly identified in the table.}
%\todo[inline]{FSM: There is an important difference between the
%  approach in \cite{sanchez08b} and the one in this paper: we do not
%  have to face the problem of free rides. Should this be mentioned
%  here?}

\begin{table*}
 \begin{center}
  \begin{tabular}{|c|l|r|}
    \hline
       & \textbf{Sentence} & $p(g_i|S)$ \\
    \hline % Los pescados grandes peque\~{n}o come
     \(S_1\) & \emph{Arrain handiak txikia jaten du}  &  \\
                        $\tau(g_{11},S_1)$ & El pez grande se come el peque\~{n}o &  0.830 \\
                        $\tau(g_{12},S_1)$ & El pescado grande se come el peque\~{n}o & 0.134 \\
                        $\tau(g_{13},S_1)$ & El pescado capaz se come el peque\~{n}o & 0.034\\
                        $\tau(g_{14},S_1)$ & El pez capaz se come el peque\~{n}o  & 0.001\\
    \hline % El padre el pescado nos ha preparado para cenar
     \(S_2\) & \emph{Aitak arraina prestatu digu afaltzeko} &  \\
                        $\tau(g_{21},S_2)$ & Padre nos ha preparado pescado para cenar & 0.846 \\
                        $\tau(g_{22},S_2)$ & Pap\'{a} nos ha preparado pescado para cenar & 0.134 \\
                        $\tau(g_{23},S_2)$ & Padre nos ha preparado pez para cenar & 0.015 \\
                        $\tau(g_{24},S_2)$ & Pap\'{a} nos ha preparado pez para cenar & 0.005 \\
    \hline % Aqu\'{\i} el pescado muy es dulce
    $S_3$  & \emph{Hemen arraina oso goxoa da}  & \\
                        $\tau(g_{31},S_3)$ & Aqu\'{\i} el pescado es muy rico & 0.912 \\
                        $\tau(g_{32},S_3)$ & Aqu\'{\i} el pescado es muy suave & 0.067 \\
                        $\tau(g_{33},S_3)$ & Aqu\'{\i} el pez es muy rico & 0.011 \\
                        $\tau(g_{34},S_3)$ & Aqu\'{\i} el pez es muy suave & 0.008 \\
                        $\tau(g_{35},S_3)$ & Aqu\'{\i} el pez es muy dulce & 0.001 \\
                        $\tau(g_{36},S_3)$ & Aqu\'{\i} el pescado es muy dulce & 0.001 \\
    \hline % Un pescado grande del mar es
    \(S_4\) & \emph{Itsasoko arrain handi bat da} & \\
                        $\tau(g_{41},S_4)$ & Es un pez grande del mar & 0.595 \\
                        $\tau(g_{42},S_4)$ & Es un pescado grande del mar & 0.403 \\
                        $\tau(g_{43},S_4)$ & Es un pez capaz del mar & 0.001 \\
                        $\tau(g_{44},S_4)$ & Es un pescado capaz del mar & 0.001 \\
    \hline % El pescado grande por esos serv\'{\i}an para alimentar
    $S_5$ &  \emph{Arrain handiak horiez baliatzen ziren elikatzeko} &  \\
                        $\tau(g_{51},S_5)$ & Los peces grandes se nutr\'{\i}an de esos & 0.862 \\
                        $\tau(g_{52},S_5)$ & Los pescados grandes se nutr\'{\i}an de esos & 0.121 \\
                        $\tau(g_{53},S_5)$ & Los peces grandes se alimentaban de esos & 0.012 \\
                        $\tau(g_{54},S_5)$ & Los peces capaces se alimentaban de esos & 0.001 \\
                        $\tau(g_{55},S_5)$ & Los pescados grandes se alimentaban de esos & 0.001 \\
                        $\tau(g_{56},S_5)$ & Los pescados capaces se alimentaban de esos & 0.001 \\
                        $\tau(g_{57},S_6)$ & Los peces capaces se nutr\'{\i}an de esos & 0.001 \\
                        $\tau(g_{58},S_7)$ & Los pescados capaces se nutr\'{\i}an de esos & 0.001 \\
    \hline

  \end{tabular}
  \caption{A sample of source-language sentences, the output of the
    possible lexical-selection paths after scoring on a
    target-language model. The scores are normalised as fractional
    counts, which are used as inputs to the maximum-entropy training
    process. In all sentences, the resultant translation with the highest share of the 
    probability mass is the most adequate.}
   \label{tb:workedexample}
 \end{center}
\end{table*}

@@

\subsubsection{Confidence intervals}
Confidence intervals for both metrics will be calculated through the
\emph{bootstrap resampling} method~\citep{efron94} as described
by~\citet{koehn04}.%\footnote{Broadly speaking, % and \citet{zhang04}
%  this involves iteratively taking a random sample (with replacements)
%  from the test set and computing the score for that sample. The
%  scores are sorted and, presuming we want the 95\% confidence
%  interval, the top- and bottom-2.5\% are removed. The highest and
%  lowest remaining are given as the interval. For more detail, refer
%  to the references provided.} 
In all cases, bootstrap resampling will
be carried out for 1,000 iterations. Where the $p = 0.05$ confidence
intervals overlap, we will also perform paired bootstrap resampling
\citep{koehn04}.

@@

There are a number of approaches to creating evaluation corpora for
lexical selection in the literature. \cite{vickrey05} use a parallel
corpus to make annotated test and training sets for experiments in
lexical selection applied to a simplified translation problem in
SMT. They use word alignments from {\sc giza++} \citep{och03a} to
annotate source language words with their translations from the
reference translation in the parallel corpus. One disadvantage of this
method is that only one translation is annotated per source language
word, meaning that accuracies may be lower because of missing
translations --- this happens when the system chooses a translation
which is adequate, but is not found in the reference translation. A
second disadvantage is that the word alignments may not be 100\%
reliable, which decreases the accuracy of the annotated corpus. An
alternative method is described by \cite{zinovjeva2000}, who manually
tags a selection of ambiguous words in English sentences with their
translation in Swedish. The size of the annotated corpus is not
mentioned. \cite{specia2005a} use a parallel corpus, but eschew using
word alignment in favour of using a bilingual dictionary and heuristic
rules based on part-of-speech and relative position. The corpus was
reviewed manually before being used in further experiments
\citep{specia2005}.

>>>>

As our method is unsupervised, and therefore the occurrences of
specific lexical selection events \((s,t,c)\) cannot be counted, a
target language model \(P_{TL}(\cdot)\) is used to compute a value for
the fractional count for disambiguation path \(g_i\), $p(g_i|S)$ after
suitable normalisation:
\begin{equation}
 \label{eq:normalising}
   p(g_i|s) = \frac{P_{TL}(\tau(g_i,S))}{\sum_{g_i \in G} P_{TL}(\tau(g_i,S))}
\end{equation}
~\\
The maximum-entropy model is trained instead using the fractional count
$p(g_i|S)$ for the events $(s,t,c)$ found in \(g_i\), that is, when in
\(g_i\) $t$ is the translation for \(s\) in context \(c\). That is, as
if event $(s,t,c)$ had been seen a fractional number $p(g_i,s)$ of
times. We prune features occurring less than a certain number of times 
in the corpus, using a development corpus to guide pruning (see section~\ref{sec:eval-ref-results}).
%@@@ \comment{Reword this to avoid repetition} 
The method used here
for lexical selection is analogous to the method used by
\cite{sanchez08b} to train a hidden-Markov-model-based part-of-speech
tagger in a rule-based machine translation system. 
%Table~\ref{tb:workedexample} shows how a collection of samples
%\(\mathcal{G}\) for the translation of Basque sentences containing the
%word \emph{arrain}.\footnote{The word \emph{arrain} `fish' has two translations
%in Spanish, the first \emph{pez} refers to the living animal, whereas
%\emph{pescado} refers to the animal as food.} 
%\comment{FSM: The correct translations
%  should be clearly identified in the table.}
%\todo[inline]{FSM: There is an important difference between the
%  approach in \cite{sanchez08b} and the one in this paper: we do not
%  have to face the problem of free rides. Should this be mentioned
%  here?}
% 
% \begin{table*}
%  \begin{center}
%   \begin{tabular}{|c|l|r|}
%     \hline
%        & \textbf{Sentence} & $p(g_i|S)$ \\
%     \hline % Los pescados grandes peque\~{n}o come
%      \(S_1\) & \emph{Arrain handiak txikia jaten du}  &  \\
%                         $\tau(g_{11},S_1)$ & El pez grande se come el peque\~{n}o &  0.830 \\
%                         $\tau(g_{12},S_1)$ & El pescado grande se come el peque\~{n}o & 0.134 \\
%                         $\tau(g_{13},S_1)$ & El pescado capaz se come el peque\~{n}o & 0.034\\
%                         $\tau(g_{14},S_1)$ & El pez capaz se come el peque\~{n}o  & 0.001\\
%     \hline % El padre el pescado nos ha preparado para cenar
%      \(S_2\) & \emph{Aitak arraina prestatu digu afaltzeko} &  \\
%                         $\tau(g_{21},S_2)$ & Padre nos ha preparado pescado para cenar & 0.846 \\
%                         $\tau(g_{22},S_2)$ & Pap\'{a} nos ha preparado pescado para cenar & 0.134 \\
%                         $\tau(g_{23},S_2)$ & Padre nos ha preparado pez para cenar & 0.015 \\
%                         $\tau(g_{24},S_2)$ & Pap\'{a} nos ha preparado pez para cenar & 0.005 \\
%     \hline % Aqu\'{\i} el pescado muy es dulce
%     $S_3$  & \emph{Hemen arraina oso goxoa da}  & \\
%                         $\tau(g_{31},S_3)$ & Aqu\'{\i} el pescado es muy rico & 0.912 \\
%                         $\tau(g_{32},S_3)$ & Aqu\'{\i} el pescado es muy suave & 0.067 \\
%                         $\tau(g_{33},S_3)$ & Aqu\'{\i} el pez es muy rico & 0.011 \\
%                         $\tau(g_{34},S_3)$ & Aqu\'{\i} el pez es muy suave & 0.008 \\
%                         $\tau(g_{35},S_3)$ & Aqu\'{\i} el pez es muy dulce & 0.001 \\
%                         $\tau(g_{36},S_3)$ & Aqu\'{\i} el pescado es muy dulce & 0.001 \\
%     \hline % Un pescado grande del mar es
%     \(S_4\) & \emph{Itsasoko arrain handi bat da} & \\
%                         $\tau(g_{41},S_4)$ & Es un pez grande del mar & 0.595 \\
%                         $\tau(g_{42},S_4)$ & Es un pescado grande del mar & 0.403 \\
%                         $\tau(g_{43},S_4)$ & Es un pez capaz del mar & 0.001 \\
%                         $\tau(g_{44},S_4)$ & Es un pescado capaz del mar & 0.001 \\
%     \hline % El pescado grande por esos serv\'{\i}an para alimentar
%     $S_5$ &  \emph{Arrain handiak horiez baliatzen ziren elikatzeko} &  \\
%                         $\tau(g_{51},S_5)$ & Los peces grandes se nutr\'{\i}an de esos & 0.862 \\
%                         $\tau(g_{52},S_5)$ & Los pescados grandes se nutr\'{\i}an de esos & 0.121 \\
%                         $\tau(g_{53},S_5)$ & Los peces grandes se alimentaban de esos & 0.012 \\
%                         $\tau(g_{54},S_5)$ & Los peces capaces se alimentaban de esos & 0.001 \\
%                         $\tau(g_{55},S_5)$ & Los pescados grandes se alimentaban de esos & 0.001 \\
%                         $\tau(g_{56},S_5)$ & Los pescados capaces se alimentaban de esos & 0.001 \\
%                         $\tau(g_{57},S_6)$ & Los peces capaces se nutr\'{\i}an de esos & 0.001 \\
%                         $\tau(g_{58},S_7)$ & Los pescados capaces se nutr\'{\i}an de esos & 0.001 \\
%     \hline
%   
%   \end{tabular}
%   \caption{A sample of source-language sentences, the output of the
%     possible lexical-selection paths after scoring on a
%     target-language model. The scores are normalised as fractional
%     counts, which are used as inputs to the maximum-entropy training
%     process. In all sentences, the resultant translation with the highest share of the 
%     probability mass is the most adequate.}
%    \label{tb:workedexample}
%  \end{center}
% \end{table*}

@@

\subsubsection{Confidence intervals}
Confidence intervals for both metrics will be calculated through the
\emph{bootstrap resampling} method~\citep{efron94} as described
by~\citet{koehn04}. %\footnote{Broadly speaking, % and \citet{zhang04}
%  this involves iteratively taking a random sample (with replacements)
%  from the test set and computing the score for that sample. The
%  scores are sorted and, presuming we want the 95\% confidence
%  interval, the top- and bottom-2.5\% are removed. The highest and
%  lowest remaining are given as the interval. For more detail, refer
%  to the references provided.} 
In all cases, bootstrap resampling will
be carried out for 1,000 iterations. Where the $p = 0.05$ confidence
intervals overlap, we will also perform paired bootstrap resampling
\citep{koehn04}.

@@

There are a number of approaches to creating evaluation corpora for
lexical selection in the literature. \cite{vickrey05} use a parallel
corpus to make annotated test and training sets for experiments in
lexical selection applied to a simplified translation problem in
SMT. They use word alignments from {\sc giza++} \citep{och03a} to
annotate source language words with their translations from the
reference translation in the parallel corpus. One disadvantage of this
method is that only one translation is annotated per source language
word, meaning that accuracies may be lower because of missing
translations --- this happens when the system chooses a translation
which is adequate, but is not found in the reference translation. A
second disadvantage is that the word alignments may not be 100\%
reliable, which decreases the accuracy of the annotated corpus. An
alternative method is described by \cite{zinovjeva2000}, who manually
tags a selection of ambiguous words in English sentences with their
translation in Swedish. The size of the annotated corpus is not
mentioned. 
%\cite{specia2005a} use a parallel corpus, but eschew using
%word alignment in favour of using a bilingual dictionary and heuristic
%rules based on part-of-speech and relative position. The corpus was
%reviewed manually before being used in further experiments
%\citep{specia2005}.
